---
title: "PipeMaster: PonAbe Worked Example"
subtitle: "Model Selection, then Parameter Estimation, from RAD-seq Data"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", dev = "pdf"
)
options(width = 90)
```

```{r cache-note, include = FALSE}
# knitr's cache is deliberately NOT used here. Trained torch objects hold
# external pointers, which do not survive the saveRDS round-trip a cache
# entry performs -- restoring one yields stale pointers and an "external
# pointer is not valid" error downstream. Instead the expensive steps below
# write to disk and are skipped on re-knit via file-existence guards, using
# PipeMaster's own save/load functions, which rebuild the pointers correctly.
```

This document runs the complete PipeMaster pipeline on the **PonAbe_TwoSpecies**
example data shipped with the package — Bornean (*Pongo pygmaeus*) and Sumatran
(*Pongo abelii*) orangutans. Nothing is downloaded, and every figure below is
produced by the code shown above it.

The order matters. We ask **which model** first, and only then **what its
parameters are**. Estimating parameters under a model the data reject gives
precise, confident, meaningless numbers.

# 0. Install or update PipeMaster

**Do this first, even if you already have PipeMaster installed.** This
tutorial relies on recent fixes to `get.data.structure()`,
`observed.sumstats()` and `nn.predict.classify()`. On an older installation it
does not merely give different numbers -- it stops with an error in section
1.2 and again in section 2.6.

```{r install, eval = FALSE}
install.packages("devtools")
devtools::install_github("gehara/PipeMaster")
```

Restart R after installing, then load it:

```{r libs}
library(PipeMaster)
set.seed(42)

OUT <- "ponabe_output"
dir.create(OUT, showWarnings = FALSE)
```

PipeMaster pulls in `ape`, `e1071`, `msm` and `torch` automatically. `torch`
downloads its own C++ backend the first time it is used, which needs a working
internet connection and a few minutes -- do it before the session rather than
during it:

```{r torch-check, eval = FALSE}
torch::torch_is_installed()   # should be TRUE
```

Two optional packages are worth having: `RANN` speeds up the
nearest-neighbour searches in the OOD diagnostics, and `bigmemory` reduces
memory use when training with several cores.

```{r optional-deps, eval = FALSE}
install.packages(c("RANN", "bigmemory"))
```

The first code chunk that touches the data will confirm your installation is
current: if `get.data.structure()` asks you for a `chrom.sizes` argument, your
copy of PipeMaster predates the fix and you need to reinstall.

# 1. Setup

## 1.1 The data

Twenty diploid individuals, ten per species, typed at 10,000 RAD loci of
100 bp each.

```{r load-data}
extdir <- system.file("extdata", "PonAbe_TwoSpecies", package = "PipeMaster")
list.files(extdir)

vcf_file   <- file.path(extdir, "PonAbe_TwoSpecies_RAD.vcf")
pop_assign <- read.table(file.path(extdir, "pop_assign.txt"), header = TRUE)
table(pop_assign$pop)
```

## 1.2 The model, attached to the data

`model_informative.RData` ships configured for whole genomes (10,000 loci
x 100 kb). `get.data.structure()` re-shapes it onto the RAD data by reading
the `##contig` header of the VCF.

```{r attach}
load(file.path(extdir, "model_informative.RData"))
model_IM <- PonAbe_IM_WGS_informative
model_IM$labels$name <- "IM"

model_IM <- get.data.structure(model_IM, path.to.vcf = vcf_file,
                               pop.assign = pop_assign)

cat(sprintf("%d loci x %s bp, %s haplotypes\n",
            nrow(model_IM$loci), model_IM$loci[1, 2],
            paste(model_IM$I[1, 4:5], collapse = " + ")))
```

**What `get.data.structure()` does.** A model carries its priors *and* a
description of the data it applies to: how many loci, how long each is, and
how many haplotypes per population. This rewrites that description
(`model$loci` and `model$I`) to match a real dataset, leaving every prior
untouched. For VCF input it reads the locus list and lengths from the
`##contig=<ID=...,length=...>` header; for PHYLIP it takes them from the
alignments. Simulations must have the same locus structure as the data, or
the simulated and observed statistics are not comparable.

Of those 10,000 loci, only 5,533 carry a variant. The other 4,467 are
**kept**: a RAD locus that came back monomorphic is evidence of low
diversity, and it belongs in the per-locus denominator of every summary
statistic. Dropping it inflates $\pi$, $\theta_W$ and Fst alike.

## 1.3 Observed summary statistics

```{r observed}
obs <- observed.sumstats(model_IM, path.to.vcf = vcf_file,
                         pop.assign = pop_assign, verbose = FALSE)

round(obs[1, c("s_mean_pi", "s_mean_thetaW", "s_mean_Fst", "s_mean_segs")], 5)
```

**What `observed.sumstats()` does.** It converts the VCF to internal
haplotype matrices and computes summary statistics with the *same C code the
simulator uses*. That shared implementation is the point: observed and
simulated statistics must be computed identically, or the network learns the
difference between the two pipelines rather than the difference between
demographic histories.

It returns two blocks of numbers, 545 in total here:

- **104 summary statistics.** Each statistic is computed per locus, then
  summarised across loci by its first four moments -- hence the `s_mean_`,
  `s_var_`, `s_skew_` and `s_kurt_` prefixes. The variance and shape terms
  carry real information: two histories can share a mean $\pi$ while
  differing in how variable $\pi$ is across the genome. Suffixes mark the
  population: `_1` Bornean, `_2` Sumatran, bare names pooled, `_1_2`
  pairwise. The underlying quantities are `segs` (segregating sites), `pi`
  (nucleotide diversity), `thetaW` (Watterson's $\theta$), `tajd` (Tajima's
  D), `ZnS` (linkage disequilibrium), `Fst`, plus `shared`, `private` and
  `fixed` differences between the pair.
- **441 folded SFS bins** -- the joint site frequency spectrum, 21 x 21 for
  20 + 20 haplotypes. Folded because the ancestral allele is unknown.

These are values per locus averaged over all 10,000 loci, which is why the
invariant loci in the previous section matter so much: they are 4,467 zeros
in the denominator.

# 2. Part I --- Which model?

The biological question is whether Bornean and Sumatran orangutans exchanged
genes after they split, or diverged in complete isolation. We build both
models and simulate under each, then answer the question twice: first with
plain ABC rejection, which has no moving parts, and then with a neural
classifier. Doing it in that order shows what the network buys you, and why.

## 2.1 The competing model

Strict isolation is the IM model with migration deleted. Everything else --
priors, divergence time, locus structure -- is identical, so the comparison
isolates exactly one factor.

```{r build-iso}
model_ISO <- model_IM
model_ISO$flags$m <- NULL
model_ISO$labels$name <- "ISO"

# confirm the -m flag really disappeared from the ms command
c(IM  = grepl(" -m ", PipeMaster:::msABC.commander(model_IM,  arg = 1)[[1]]),
  ISO = grepl(" -m ", PipeMaster:::msABC.commander(model_ISO, arg = 1)[[1]]))
```

## 2.2 Reference tables

**What `sim.sumstats()` does.** For each simulation it (1) draws one value
per free parameter from that parameter's prior, honouring the model's
conditions, (2) converts them to coalescent units and builds an `ms` command,
(3) simulates genealogies and mutations for all 10,000 loci with the built-in
msABC engine, and (4) computes the same 545 statistics as
`observed.sumstats()`. One row of the reference table = one parameter draw
plus the statistics it produced.

This is the "reference table" that everything downstream consumes: the
classifier learns which model produces which statistics, and the regression
network learns which parameter values produce which statistics. Simulation is
the only place the demographic model actually enters -- every later step sees
only this table.

`batch.size` controls how many simulations go to each C call (memory and
progress granularity, not the total); `ncores` sets independent worker
processes. Use `sim.scrm.sumstats()` instead for long loci with
recombination, where the scrm engine is far faster.

Equal numbers per model: the classifier learns whatever class balance you
give it, so an unbalanced table biases the model probabilities directly.

```{r simulate-show, eval = FALSE}
NSIMS <- 2000
CORES <- 8

sim.sumstats(model_IM,  nsims = NSIMS, batch.size = 8, mu.rates = 2e-8,
             ncores = CORES, path = OUT, output.name = "IM")

sim.sumstats(model_ISO, nsims = NSIMS, batch.size = 8, mu.rates = 2e-8,
             ncores = CORES, path = OUT, output.name = "ISO")
```

```{r simulate-run, include = FALSE}
NSIMS <- 2000
CORES <- 8

for (nm in c("IM", "ISO")) {
  if (file.exists(file.path(OUT, paste0("SIMS_", nm, ".txt")))) next
  m <- if (nm == "IM") model_IM else model_ISO
  sim.sumstats(m, nsims = NSIMS, batch.size = 8, mu.rates = 2e-8,
               ncores = CORES, path = OUT, output.name = nm)
}
```

```{r read-reftables}
rt_IM  <- read.table(file.path(OUT, "SIMS_IM.txt"),  header = TRUE)
rt_ISO <- read.table(file.path(OUT, "SIMS_ISO.txt"), header = TRUE)

param_cols <- c("Ne0.pop1", "Ne0.pop2", "Ne1.pop1", "Ne1.pop2",
                "join1", "mig0.1_2", "mig0.2_1")
derived    <- c("Ne.anc_1_2", "t.Ne1.pop1", "t.Ne1.pop2", "t.Ne.anc_1_2")
nuisance   <- c("mean.rate", "sd.rate")
stat_cols  <- setdiff(colnames(rt_IM), c(param_cols, derived, nuisance))

cat(sprintf("IM  %d x %d\nISO %d x %d\n%d statistics\n",
            nrow(rt_IM), ncol(rt_IM), nrow(rt_ISO), ncol(rt_ISO),
            length(stat_cols)))

obs_raw <- setNames(as.numeric(obs[1, stat_cols]), stat_cols)
```

The ISO table has two fewer columns -- no `mig0.1_2`, no `mig0.2_1` -- so the
two are stacked on the statistics they share, plus a label:

```{r stack}
rt_both <- rbind(
  data.frame(model = "IM",  rt_IM[,  stat_cols], check.names = FALSE),
  data.frame(model = "ISO", rt_ISO[, stat_cols], check.names = FALSE))

table(rt_both$model)
```

## 2.3 Can the observation be produced by either model?

Run this **before** training. It asks whether the observed data lie inside
each model's prior-predictive support.

```{r ood-pretrain-classify, fig.width = 11, fig.height = 11}
pre_c <- OOD.pretrain.classify(reftable = rt_both, model_col = "model",
                               observed = obs_raw, stat.cols = stat_cols,
                               plot = TRUE, verbose = FALSE)

cat("verdict:", pre_c$verdict, "\n")
cat("universal outliers:", sum(pre_c$universal_mask), "of", length(stat_cols), "\n")
```

**What `OOD.pretrain.classify()` does.** The multi-model counterpart of
`OOD.pretrain()`. It runs the per-statistic support check **separately for
each candidate model**, then projects every model's simulations plus the
observation into one shared PCA and one shared **PLS-DA** (discriminant
analysis: components chosen to separate the classes, rather than to track
parameters as in the single-model case). The scatter plots show whether the
two models occupy distinct regions at all, and where the observation falls
relative to both.

The critical output is `universal_mask` -- statistics whose observed value
falls outside *every* candidate model's support. No model explains them, so
they cannot discriminate between models and are noise to the classifier.
Statistics rejected by only *some* models are the informative ones. A large
universal fraction fails the check, and the honest response is a better
model, not a better classifier.

## 2.4 First, the simple way: model choice by pure rejection

Before reaching for a neural network, do it the plain way. ABC rejection has
no training, no hyperparameters and nothing to go wrong: measure how far each
simulation is from the observation, keep the closest few, and count how many
came from each model.

`abc.rejection()` does the work. It is internal to the package, so it needs
the `:::` prefix. It expects parameter columns, so we code the model label as
a number and let it ride along: whatever it accepts, the labels come back
attached.

```{r rejection-modelchoice}
rt_rej <- rbind(
  data.frame(model = 1, rt_IM[,  stat_cols], check.names = FALSE),   # 1 = IM
  data.frame(model = 2, rt_ISO[, stat_cols], check.names = FALSE))   # 2 = ISO

ms_rej <- PipeMaster:::abc.rejection(
  reftable   = rt_rej,
  observed   = obs_raw,
  param.cols = "model",
  tol        = 0.01,
  distance   = "sd")

counts <- table(factor(c("IM", "ISO")[ms_rej$abc[, "model"]],
                       levels = c("IM", "ISO")))
counts
round(prop.table(counts), 3)          # posterior model probabilities
```

**What `abc.rejection()` does.** It scales every statistic so they contribute
comparably -- by standard deviation for `distance = "sd"`, by median absolute
deviation for `"mad"`, or using the full covariance for `"mahalanobis"` --
computes the Euclidean distance from each simulation to the observation, and
keeps the closest `tol` fraction. Statistics that never vary get a scale of 1
rather than being dropped, so they contribute nothing when the observation
matches them. There is no training and nothing to tune beyond `tol`.

With equal simulations per model and no prior preference between them, the
**proportion of accepted simulations from each model is its posterior
probability**, and their ratio is the **Bayes factor**.

```{r rejection-bf}
cat(sprintf("Bayes factor IM/ISO = %.2f\n", counts[["IM"]] / counts[["ISO"]]))
```

Two things are worth pausing on.

**The answer is weak.** The accepted set is far from one-sided, so the Bayes
factor sits near 1 -- barely worth mentioning by any of the usual scales. Yet
the classifier in the next section separates these same two models almost
perfectly. Both are computed from the same simulations, so the difference is
not information in the data; it is how the two methods use it.

**Look at the distance range printed above.** Even the accepted simulations --
the closest 1% of `r format(nrow(rt_rej), big.mark = ",")` -- sit at distances
of roughly 77 to 83, measured in standard deviations of the statistics. The
best match is not a good match. Nothing stands out from the crowd, because
across `r length(stat_cols)` statistics every simulation is far from the
observation in *some* direction, and Euclidean distance adds all those
directions up with equal weight. This is the **curse of dimensionality**, and
it is what blunts plain rejection: only a handful of these statistics respond
to migration, and the rest bury that signal in noise. "Closest" stops meaning
"most similar in the ways that matter".

The standard remedies are to hand-pick a few informative statistics, or to
reduce dimension first with PLS (`pls = TRUE`, which projects onto components
chosen for their relationship to the parameters). The neural network is a
third answer: rather than being told which statistics matter, it **learns**
the weighting from the simulations. That is the entire reason for the next
section.

## 2.5 Train the classifier

```{r train-classifier-show, eval = FALSE}
clf <- tune.nn.classify(rt_both, model_col = "model",
                        max_epochs = 40L, eta = 3L,
                        n_searches = 2L, cores = 2L, gpus = 0L,
                        seed = 42, verbose = FALSE)

save.classifier(clf, file.path(OUT, "clf"))
```

```{r train-classifier-run, include = FALSE}
clf_path <- file.path(OUT, "clf")
if (file.exists(clf_path)) {
  clf <- load.classifier(clf_path)
} else {
  clf <- tune.nn.classify(rt_both, model_col = "model",
                          max_epochs = 40L, eta = 3L,
                          n_searches = 2L, cores = 2L, gpus = 0L,
                          seed = 42, verbose = FALSE)
  save.classifier(clf, clf_path)
}
```

```{r train-classifier-out}
cat("validation accuracy:", round(clf$metrics$accuracy, 4), "\n")
clf$metrics$confusion
```

**What `tune.nn.classify()` does.** Same architecture and same Hyperband
search as `tune.nn()`, but the target is a class label rather than a
parameter vector, so it minimises **cross-entropy** instead of squared error.
Features are augmented and z-scored identically. Parameter columns are
detected and excluded automatically -- the classifier must decide from
statistics alone, since the observed data have no parameters attached. The
validation split is **stratified by class**, so both models are represented
in the held-out set.

**Reading the output.** `best_val_loss` is validation cross-entropy (lower is
better; $\ln 2 \approx 0.69$ is the value of a coin flip for two classes).
The confusion matrix has true classes in rows and predicted in columns, so
off-diagonal entries are the simulations the classifier got wrong -- and their
direction tells you *which* model gets mistaken for which.

Report this number. High accuracy means the two models are **distinguishable
given this data structure** -- a precondition for the comparison meaning
anything. If accuracy sat near 0.5, the models would not be identifiable from
these statistics and no probability computed afterwards would be worth
reporting.

Accuracy of exactly 1.0 deserves a moment's suspicion rather than
satisfaction: with only 2,000 simulations per class, a validation set of ~100
simulations can be separated perfectly by a model that would still fail on
fresh data. Treat it as "clearly separable", not as a precise number.

`save.classifier()` and `load.classifier()` write and rebuild the trained
network. Use them rather than `saveRDS()`: torch objects hold external
pointers that do not survive a plain serialisation round-trip, and restoring
one gives an `external pointer is not valid` error.

## 2.6 Classify the observation

Three output rules, answering slightly different questions:

```{r classify}
res <- lapply(c("softmax", "mahalanobis", "latent_nn"), function(m)
  nn.predict.classify(clf, observed = obs_raw, method = m,
                      reftable = rt_both, model_col = "model", verbose = FALSE))

do.call(rbind, lapply(seq_along(res), function(i) data.frame(
  method = c("softmax", "mahalanobis", "latent_nn")[i],
  best   = res[[i]]$best_model,
  P_IM   = round(res[[i]]$prob_mean[["IM"]],  4),
  P_ISO  = round(res[[i]]$prob_mean[["ISO"]], 4))))
```

- **softmax** -- the classifier's native output. Measures which class
  *direction* the observation aligns with. It always names a winner, even for
  an observation lying far outside both classes.
- **mahalanobis** -- distance to each class centroid under the pooled
  within-class covariance. Density-aware; the one to trust when the
  observation may be out of distribution.
- **latent_nn** -- distance to the nearest simulation of each class. Purely
  geometric.

Agreement across all three is the result you want. Disagreement means the
observation sits where the classifier is extrapolating.

## 2.7 Is the observation inside the class it was assigned?

A model probability is only meaningful if the observation actually lies within
the winning class's learned manifold.

```{r ood-posttrain-classify, fig.width = 11, fig.height = 7.5}
post_c <- OOD.posttrain.classify(clf, observed = obs_raw, reftable = rt_both,
                                 model_col = "model", pretrain = pre_c,
                                 plot = TRUE, verbose = FALSE)

cat("verdict:        ", post_c$verdict, "\n")
cat("predicted class:", post_c$predicted_class, "\n")
cat("margin:         ", round(post_c$margin, 4), "\n")
```

**What `OOD.posttrain.classify()` does.** It takes the classifier's
penultimate-layer representation -- the space in which it separates the
models -- and asks, for each class, whether the observation lies inside that
class's cloud of simulations, using the same nearest-neighbour test against a
leave-one-out null. It then cross-references the answer with the predicted
class.

This is the check that makes a model probability meaningful. A softmax always
sums to one and will hand you 0.99 for an observation sitting nowhere near
either class: the classifier reports which side of its decision boundary you
are on, and that boundary extends to infinity. Asking separately whether you
are *near any training data* is what distinguishes "the data favour IM" from
"IM is the closer of two models that both fail".

`$margin` is $P(\text{top 1}) - P(\text{top 2})$, a one-number measure of how
decisive the call was. A high margin with a `fail` verdict is the dangerous
combination -- confident and unfounded.

| Verdict | Meaning |
|---|---|
| `pass` | Observation lies inside the predicted class. Probability is usable. |
| `warn` | Inside *some* class, but not the predicted one. Prediction unreliable. |
| `fail` | Outside every class. The classifier is extrapolating; the probability is meaningless. |

With real data `fail` is common and important: it means none of your candidate
models can produce the observed data, so the "winner" is merely the least-bad
of a bad set. Add a model rather than report the ranking. When that happens,
`OOD.attribute.classify()` attributes the out-of-distribution-ness to
individual statistics via integrated gradients.

# 3. Part II --- Parameter estimation under the selected model

Only now, having established that the data prefer IM *and* that the
observation lies inside the IM manifold, do we estimate parameters.

## 3.1 Which parameters are free?

Four of the eleven parameter columns are not free -- they are deterministic
consequences of the model's conditions. Feeding them to the network asks it to
learn duplicates.

```{r free-params}
c(times_equal_join = all(rt_IM$t.Ne1.pop1   == rt_IM$join1),
  anc_is_sum       = all(rt_IM$Ne.anc_1_2   == rt_IM$Ne1.pop1 + rt_IM$Ne1.pop2))

param_cols
```

`t.Ne1.pop1`, `t.Ne1.pop2` and `t.Ne.anc_1_2` all equal `join1` through the
model's `=` time conditions; `Ne.anc_1_2` is the sum of the daughter sizes
because `model$sum_anc_ne` is `TRUE`. Verify this for your own model rather
than assuming it -- the conditions are model-specific.

## 3.2 First, the simple way: estimation by pure rejection

Same principle as the model-choice rejection above, but now we keep the
*parameter values* of the accepted simulations instead of counting their
labels. `abc.rejection()` does this directly. It is internal, so it needs the
`:::` prefix.

```{r rejection-params}
post_rej <- PipeMaster:::abc.rejection(
  reftable   = rt_IM[, c(param_cols, stat_cols)],
  observed   = obs_raw,
  param.cols = param_cols,
  tol        = 0.05,
  distance   = "sd")

summary(post_rej)
```

**What `abc.rejection()` does.** Scales each statistic (by its standard
deviation for `distance = "sd"`, by median absolute deviation for `"mad"`, or
using the full covariance for `"mahalanobis"`), computes Euclidean distance
from every simulation to the observation, and keeps the closest `tol`
fraction. The accepted rows' parameter values *are* the posterior sample --
no adjustment, no model fitting. The point estimate is their distance-weighted
mean. Setting `pls = TRUE` projects the statistics onto PLS components first,
which is the standard fix for the dimensionality problem seen above.

```{r plot-rejection, fig.width = 10, fig.height = 8}
plot(post_rej, bw.adjust = 2)
```

**This is the honest baseline, and it has one guarantee the network cannot
offer**: every value in the posterior is a parameter that was actually
simulated, so the result can never leave the prior. Compare these estimates
with the network's in section 3.5 -- where the network wanders outside the
prior, rejection by construction cannot.

The cost is resolution. With `r nrow(rt_IM)` simulations and `tol = 0.05`,
the posterior rests on roughly 100 accepted simulations, and the same
dimensionality problem from section 2.4 applies: distances are computed over
hundreds of statistics of which only a few are informative. Expect posteriors
that look close to the prior for parameters the statistics constrain weakly.

*(A cosmetic wrinkle: `summary()` labels this block "ABC-NN regression
posterior" because it shares the `posterior` class with `nn.predict()`
output. No network was involved here.)*

## 3.3 Prior-predictive check

```{r ood-pretrain, fig.width = 11, fig.height = 11}
pre <- OOD.pretrain(observed = obs_raw,
                    reftable = rt_IM[, c(param_cols, stat_cols)],
                    stat.cols = stat_cols, param.cols = param_cols,
                    plot = TRUE, verbose = FALSE)

cat("verdict:", pre$overall, "\n")
cat("outlier statistics:", sum(pre$percentiles$outlier),
    "of", length(stat_cols), "\n")
```

**What `OOD.pretrain()` does.** Three independent checks that the observed
data are reachable under this model and prior, all run *before* training so
you can stop early:

1. **Per-statistic support.** For each of the 545 statistics, does the
   observed value fall inside the range the simulations produced? The
   fraction outside sets a tier: pass below 10%, warn to 25%, fail above.
2. **Nearest-neighbour density in PCA space.** Statistics are correlated, so
   an observation can sit inside every marginal range yet in a hole in the
   joint distribution. This projects to principal components and compares the
   distance from the observation to its nearest simulation against the
   distribution of simulation-to-simulation distances. A p-value near zero
   means the observation is isolated.
3. **The same test in PLS space**, using components chosen to correlate with
   the parameters rather than to maximise variance. Passing `param.cols`
   enables it. PLS can expose a problem PCA misses when the parameter-relevant
   signal is a small share of total variance.

Each runs on all statistics and again on the subset that survived check 1,
because a handful of wild outliers can dominate a distance in 545 dimensions.

**Reading the output.** `$overall` combines the tiers. `$percentiles` holds
each statistic's position within the simulated distribution, and
`$percentiles$outlier` flags those outside -- inspect them with
`OOD.outliers()`, and drill into the projections with
`OOD.projection.diagnose()`.

Note the accessor: `OOD.pretrain()` and `OOD.posttrain()` report in
**`$overall`**, while the classifier variants above use **`$verdict`**.

## 3.4 Train the regression network

```{r train-nn-show, eval = FALSE}
tune_result <- tune.nn(rt_IM, param.cols = param_cols, type = "sumstat",
                       exclude.cols = c(derived, nuisance),
                       max_epochs = 40L, eta = 3L,
                       n_searches = 2L, top_k = 2L,
                       cores = 2L, gpus = 0L, seed = 42, verbose = FALSE)

save.tune.result(tune_result, file.path(OUT, "tune_IM"))
```

```{r train-nn-run, include = FALSE}
tune_path <- file.path(OUT, "tune_IM")
if (dir.exists(tune_path)) {
  tune_result <- load.tune.result(tune_path)
} else {
  tune_result <- tune.nn(rt_IM, param.cols = param_cols, type = "sumstat",
                         exclude.cols = c(derived, nuisance),
                         max_epochs = 40L, eta = 3L,
                         n_searches = 2L, top_k = 2L,
                         cores = 2L, gpus = 0L, seed = 42, verbose = FALSE)
  save.tune.result(tune_result, tune_path)
}
```

```{r train-nn-out}
cat("best validation loss:", round(tune_result$best_val_loss, 5), "\n")
```

**What `tune.nn()` does.** It learns the inverse map, statistics $\rightarrow$
parameters, by regression on the reference table. Three details matter for
reading the output:

- **Feature augmentation.** Each statistic enters twice, as
  `cbind(raw, log1p(abs(raw)))`, then z-scored. Our 545 statistics become 1090
  inputs. The log copy lets the network use both absolute and relative
  changes in quantities like $\theta_W$ that vary over orders of magnitude.
- **Targets are modelled in log space**, z-scored, and inverted with
  `exp()`. This keeps Ne and time parameters positive and makes their priors
  roughly uniform on the learned scale. **It also means predictions are not
  confined to the prior range** -- `exp()` of an extrapolated z-score can land
  anywhere. An estimate outside the prior is therefore a symptom of an
  under-trained or extrapolating network, not a bug.
- **Hyperband search.** `n_searches` independent searches each run a full
  Hyperband: many configurations get a small training budget, the worst are
  discarded, survivors get `eta`-times more, repeat. `top_k` keeps the best
  models as an ensemble rather than one winner.

`best_val_loss` is mean squared error on the held-out split, in z-scored log
units -- useful for comparing runs on the same data, not interpretable as a
parameter error. `r2` and mean percentage error in original units are in
`tune_result$metrics`.

**Ensemble prediction.** The point estimate is not a plain average of the
`top_k` models. Validation simulations are weighted by a Gaussian kernel on
their distance to the observation, each model's validation loss is recomputed
under those weights, and the models are combined by a softmax of those
*local* losses. A model that is mediocre overall but accurate in the
neighbourhood of your data is weighted up.

## 3.5 Posterior

Two uncertainty methods, one call each, merged for plotting.

```{r predict}
post_boot <- nn.predict(tune_result, observed = obs_raw, reftable = rt_IM,
                        param.cols = param_cols, method = "bootstrap",
                        n_boot = 2000L, seed = 42, verbose = FALSE)

post_abc <- nn.predict(tune_result, observed = obs_raw, reftable = rt_IM,
                       param.cols = param_cols, method = "ABC_NN_regression",
                       tolerance = 0.2, seed = 42, verbose = FALSE)

pred <- post_abc
pred$bootstrap <- post_boot$bootstrap
class(pred) <- "posterior"

summary(pred)
```

**Reading `summary()`.** *Point estimate* is the proximity-weighted ensemble
prediction described above -- a single best guess, no uncertainty attached.
*Residual bootstrap* gives the median and quantiles of the bootstrap
replicates. *ABC* gives the same for the regression-adjusted accepted
simulations. The two intervals are not interchangeable: compare the bootstrap
against itself across runs to judge network stability, and report the ABC
interval as the posterior.

Both are back-transformed to natural units -- Ne in individuals, `join1` in
generations, migration in Nm (migrants per generation) since
`model$mig_scale` is `"Nm"`.

These two are not two ways of computing the same thing. They answer different
questions, and it is worth being precise about how each works.

**What the residual bootstrap does.** It asks: *how wrong is this network
usually, on simulations that look like my data?*

1. Run the trained network on the **validation** simulations -- data it never
   trained on -- and compute residuals, `true - predicted`, on the original
   parameter scale.
2. Measure the Euclidean distance from every validation simulation to the
   observed data in standardised statistic space.
3. Weight each residual by that distance through an **Epanechnikov kernel**,
   $w \propto 1 - u^2$ for $u = d/h \le 1$, with the bandwidth $h$ set to the
   10th percentile of distances. Residuals from simulations unlike your data
   get zero weight. (If fewer than 50 survive, it falls back to a Gaussian
   kernel with $h$ = median distance, so the interval never collapses.)
4. Draw `n_boot` residuals with replacement in proportion to those weights and
   add each to the point estimate.

So the interval is *centred on the point estimate* and its width is the local
error of the network. The prior never enters. This is a **prediction
interval**: it tells you how much the network's answer would wobble, not what
the data say about the parameter. It can sit entirely outside the prior.

**What ABC-NN regression does.** It asks: *which parameter values could have
generated data like mine?* This is Beaumont et al. (2002), with the network
serving as the regression function.

1. Compute $\lVert S_i - S_{obs} \rVert$ for **every row of the reference
   table** in standardised statistic space.
2. Accept the closest `tolerance` fraction -- plain ABC rejection so far.
3. The accepted rows carry their own true parameter values $\theta_i$, drawn
   from the prior. A plain ABC posterior would stop here and report them.
4. **Regression adjustment**: each accepted $\theta_i$ is shifted by how the
   network's prediction differs between that simulation and the observation,
   $$\theta_i^{adj} = \theta_i - \left(g(S_i) - g(S_{obs})\right)$$
   correcting for the fact that an accepted simulation is *near* the
   observation but not identical to it.

Because it starts from parameter values that were actually drawn from the
prior and only nudges them, the result is a **posterior**: prior-supported by
construction, and it cannot leave the prior range by more than the adjustment.

The practical consequence: if the two disagree sharply, trust the ABC
posterior about the parameter and read the bootstrap as a statement about how
much the network is struggling.

`tolerance = 0.2` keeps 20% of simulations, appropriate at this small budget.
At production scale (50,000+ simulations) drop it to 0.05.

```{r plot-posterior, fig.width = 10, fig.height = 8}
plot(pred, bw.adjust = 2)
```

One panel per free parameter. Each shows the **prior** (grey), the **residual
bootstrap** (blue) and the **ABC posterior** (steel blue), so the question
"did the data actually tell us anything?" is answered by eye: a posterior
that simply reproduces the prior means the statistics carry no information
about that parameter, and one pinned against a prior edge means the bound is
doing the work and should be widened. `bw.adjust` only smooths the density
estimate; it does not change the samples.

## 3.6 Post-training diagnostics

```{r ood-posttrain, fig.width = 11, fig.height = 7.5}
# note: OOD.posttrain() has no `verbose` argument, unlike OOD.pretrain()
post <- OOD.posttrain(trained.nn = tune_result, observed = obs_raw,
                      reftable = rt_IM[, c(param_cols, stat_cols)],
                      pretrain = pre, plot = TRUE)

cat("verdict:", post$overall, "\n")
```

**What `OOD.posttrain()` does.** It re-runs the pre-training checks (reusing
their PCA and PLS fits when you pass `pretrain =`, which is faster and gives
one combined verdict) and adds two that require a *trained* network:

4. **NN-latent density.** The same nearest-neighbour test, but in the
   penultimate layer's activation space -- the representation the network
   actually regresses from. PCA and PLS are linear projections; the network's
   internal space is curved, so an observation can look comfortable in every
   linear view and still sit in a hole in the space where prediction happens.
   This is the check that most often catches a genuine problem.
5. **Ensemble disagreement.** The `top_k` models all fit the training data
   well, so where they still disagree is where the data did not constrain
   them. This computes the coefficient of variation across the ensemble at
   the observed point and compares it to the same quantity across all
   simulations. High disagreement means the network is extrapolating exactly
   where you are asking it to predict -- a direct warning that the point
   estimate is unreliable.

The distinction from `OOD.pretrain()` is worth keeping straight: pre-training
asks whether the **model and prior** can produce your data; post-training asks
whether the **trained network** can be trusted at your data. A model can pass
the first and fail the second simply because the reference table was too
small -- which is what happens at this document's budget.

# 4. Simulation budget

Everything above runs at **2,000 simulations per model** so the document
knits in minutes. That is a demonstration budget, not an analysis budget, and
the diagnostics will say so -- expect `OOD.posttrain()` to return `warn` with
high ensemble disagreement at this size. That is the tool working correctly.

Measured throughput for this model (10,000 loci x 100 bp, 40 haplotypes):
approximately **30,600 simulations/hour on 4 cores** and **73,300/hour on 8**.

| Tier | sims / model | 4 cores | 8 cores | Use for |
|---|---|---|---|---|
| Demonstration | 2,000 | ~4 min | ~2 min | This document |
| Teaching | 10,000 | ~20 min | ~8 min | An interpretable posterior |
| Production | 50,000 | ~1.7 h | ~40 min | Results you would report |

For a 7-parameter model, 50,000 simulations is the point at which the
posterior stops moving when you add more.

# References

- Beaumont M.A., Zhang W. & Balding D.J. (2002) Approximate Bayesian
  computation in population genetics. *Genetics* 162: 2025-2035.
- Locke D.P. et al. (2011) Comparative and demographic analysis of orang-utan
  genomes. *Nature* 469: 529-533.
- Lee K. et al. (2018) A simple unified framework for detecting
  out-of-distribution samples and adversarial attacks. *NeurIPS*.
- Pavlidis P., Laurent S. & Stephan W. (2010) msABC: a modification of
  Hudson's ms to facilitate multi-locus ABC analysis. *Molecular Ecology
  Resources* 10: 723-727.
